Developing a Visual Studio Code Language Support Extension for the Snail Programming Language

### 2023-06-30 paper.tex
feat: finish content

    A high quality programming environment (often an integrated development environment, or IDE) can be vital to enhancing developer productivity. Visual Studio Code (VS Code) is a popular, open-source text editor maintained by Microsoft. VS Code delivers language-specific features through freely dowloadable, community-built extensions on an online marketplace. Many of these extensions allow developers to take advantage of editing features such as syntax highlighting, code-autocompletion, or debugging support. The snail language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming langauge meant to be implemented in a one-semester undergraduate course. We present a new VS Code extension to provide language support for the snail language. The extension implements support for syntax highlighting, rudimentary auto-completion, and static error-checking diagnostics using VS Code's Langauge Server Protocol. This report summarizes the contents of a VS Code extension and gives an overview of how an extension runs, particularly highlighting the functions of VS Code's Langauge Server Protocol. I also discuss how this extension can be futher developed to make use of VS Code's Debug Adapter Protocol to implement a debugger with breakpoints, start/stop behavior, and variable inspection.

@@

Much of software development in the present day takes place in integrated development environments (IDEs) \cite{JetBrains_2019}. An IDE is a collection of software development tools, such as a code editor, debugger, and build system, often unified under a similar user interface, with the goal of simplifying the software development process and enhancing developer productivity \cite{Gillis_Silverthorne_2018,Shyniaieva_2023}. In addition to composing these tools together, many IDEs also offer advanced features within their code editors such as syntax highlighting, code auto-completion, and error-checking diagnostics. Today, there are any number of different IDEs available for use with any given programming language, many offered as freely-downloadable for public use. For example, a developer looking to write code in Java may choose to do so in Eclipse, IntelliJ IDEA, or BlueJ \cite{EclipseFoundation_2023, JetBrains_2023,KingsProgrammingEducationToolsGroup_2022}. With so many high quality IDEs available today, it is no surprise that 75\% of software developers today use an IDE in their everyday work \cite{JetBrains_2019}. Clearly, the IDE has become an integral part of the software development process today \cite{Vaniukov_2023}.

Visual Studio Code (VS Code), is a popular, open source text editor maintained by Microsoft \cite{StackOverflow_2022,Microsoft_2023a}. When first downloaded, VS Code is a lightweight text editor with minimal features. However, a number of community-built, freely downloadable extensions offer advanced langauge features. These extensions can be dowloaded on VS Code's online extension marketplace, and can turn VS Code into a very fast ,robust, and powerful development environment for any programming task or language \cite{Microsoft_2023b}.

The snail programming language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming language meant to be implemented in a one-semester undergraduate course \cite{Angstadt_2023a}. In order to be implemented in a short time frame, snail is defined by limited features and a relatively annoying syntax. The language lacks a for-loop structure, opting instead to provide only while-loops. Each if statment \emph{requires} an else clause, even when the developer does not need to use one. Every statement must end with a semi-colon, which is not an issue until you accidentally forget one, and the resulting parse error message is wildly uninformative. While this design makes snail easier to implement, it makes it hard for a developer to write programs in snail.

@@

This report presents the Snail Language Support VS Code extension, which seeks to address the lack of programming support tools for the snail programming language. Snail Langauge Support provides several important features to make programming in the snail language easier. First, it features syntax highlighting to make reading snail code easier and help highlight key structures or keywords of the language. Further, it features rudimentary auto completion with auto-closing brackets, braces, and quotes, as well as if-else, while loop, and class definition snippets, reducing the burden of memorizing snail's strict and unintuitive syntax. The extension also has automatic, real-time error checking diagnostics that allow a user to see syntax or parse errors in a snail program before running it for themselves. Finally, the extension is structured to support a full debugger with breakpoints, step-in, step-over, and step-out functionality.

This paper will outline the process of building the Snail Language Support VS Code extension.\footnote{This paper is formatted using \LaTeX. Recognizing that using \LaTeX \hspace{1pt} is really hard, the \LaTeX \hspace{1pt} source code for this paper can be found hosted on GitHub: \url{https://github.com/CharlesReinhardt/snail-language-support-paper/}} Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for new programming langauges. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide realtime error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the role of VS Code's debug adapter protocol (DAP) \cite{Microsoft_2021a}. Finally, we will review good software development practices such as version control and documentation.

@@

While early IDEs certainly had their merits, they usually only supported one programming language. Microsoft's Visual Studio, released in 1997, was one of the first IDEs to package support for a variety of programming languages in one piece of software \cite{Microsoft_1997}. Visual Studio also featured extensive tools to aid in development for software on the early internet. 
T
oday, we see a trend towards open source IDEs such as Eclipse, NetBeans, or VS Code \cite{Apache_2023, EclipseFoundation_2023, Microsoft_2023a}. Even open source editors such as these include advanced development features such as an intelligent code editor, debugging tools, and version control integration. Many of these editors have advanced features within their code editors, providing functions such syntax highlighting, code autocompletion, and realtime error diagnostics.

There are many benefits to developing software using modern IDEs. By bundling code editing, build systems, and program execution into one tool, modern IDEs reduce the time and effort a developer needs to put forth in order to test a piece of code \cite{Gillis_Silverthorne_2018}. This also reduces the number of decisions a developer has to make while developing code, which can help increase productivity. Using an IDE can also standardize the software development process, by either helping a group of people use a consistent UI (and know how to help eachother), or allow a developer to avoid switching applications to complete a single task \cite{Veracode_2020}.

@@

Debugging is the process of searching for and fixing unexpected errors in a piece of code \cite{AWS_2023}. Early techniques of debugging software involved physically printing machine output and reading, step-by-step, through code and output, searching for a potential error \cite{RevDeBug_2020}. Some computer systems, such as the IBM 704 Data Processing system released in 2968, allowed programmers to print information stored in specified memory locations in specified forms to assist with debugging activities \cite{MIT_1957, IBMArchives_2023}.

@@

Visual Studio Code is a popular, open source code editor. Visual Studio Code is designed to be fast and lightweight, with a focus on allowing a developer to write, test, adn debug source code quickly \cite{Microsoft_2023c}. To achieve this, VS Code uses native, web, and language-specific technologoies. Tools like Electron allow VS Code to run on multiple platforms and operating systems using common web technologies like JavaScript, HTML, and CSS \cite{Electron_2023}.

@@

Most VS Code extensions share a similar directory structure. These structures may vary depending on what purpose the extension is meant to serve. In figure \ref{fig:directory-structure}, we see a diagram of a typical language support VS Code extension. First, the \lstinline{.vscode} directory configures how we run and test our exstension locally. Through the \lstinline{launch.json} and \lstinline{tasks.json} files, we define various tasks and configurations to streamline the build and testing process.\footnote{From personal experience, we strongly recommend taking the time to understand these files before developing a VS Code extension. It will likely save painstaking debugging in the near future. See launch configurations: \url{https://code.visualstudio.com/docs/editor/debugging\#_launch-configurations} and tasks: \url{https://code.visualstudio.com/docs/editor/tasks}} Note that these configuration files do not get packaged with a released VS Code extension. Next, a \lstinline{README.md} file allows us to advertise the features of our extension. This file is rendered in the VS Code extension marketplace and is meant to be a user's first impression of an extension.

@@

The \lstinline{src} directory contains the aforementioned TypeScript source code that runs our extension. In this directory, the \lstinline{extension.ts} file defines special actions for our extension and launches the extension within an existing VS Code window. The \lstinline{server.ts} file defines a language server that allows our extension to take advantage of realtime error diagnostics. We further discuss language servers later in this paper.

@@

When a VS Code window launches, it also launches all extensions the user has downloaded and enabled by calling each extension's \lstinline{activate} function, defined in the extension's source code. In Snail Language Support's case, the \lstinline{activate} function is defined in the \lstinline{extension.ts} file. This file has access to the VS Code extension API, allowing a developer to define certain VS Code client UI updates \cite{Microsoft_2023d}. For example, Snail Language Support can display a VS Code UI style error message to the user if the user does not have a compatibale version of the snail language dowloaded.

The \lstinline{activate} function is responsible for a few different behaviors. First, the \lstinline{activate} function registers a few command handlers to watch for commands input from a user. A command handler is an example of the observer design pattern, which allows a developer to address a variety of potential input events that are experienced during runtime \cite{GangOfFour_1995}. A developer can register an observer that listens for a particular event and executes a block of code once that event is detected. Snail Language Support defines a few command handlers to detect when a user chooses to run or debug a snail file. We will touch on the topic of debugging within Snail Language Support more later.

@@

Once the extension setup inside the \lstinline{activate} function is complete, VS Code enables the extension within the workspace, allowing a developer to utilize the functions of an extension. Once the VS Code window is closed, or the extension is manually disabled, VS Code calls the \lstinline{deactivate} function, which shuts down any separate processes left over from the extension run.

To inspect Snail Language Support's \lstinline{extension.ts} file more closely, see appendix \ref{app:extension-ts}

@@

First, we can define some metadata about our extension, such as a name, author, description, and version. We can also define which version of the VS Code engine this extension expects. This ensures that an extension knows which VS Code APIs it has proper access to. For example, an extension that expects a VS Code version of \lstinline{1.7.0} won't be able to take advantage of an API released in \lstinline{1.8.0} \cite{Microsoft_2023e}. We can also define some categories that our extension falls under, helping our extension gain visibility on the online VS Code marketplace.

@@

Snail Language Support also provides language configuration settings supported by a VS Code extension. These configurations are found in \lstinline{language-configuration.json} file. We can define what is considered a commend through the \lstinline{comments.lineComment} or \lstinline{comments.blockComment} attributes, which allows a user to comment lines using VS Code's comment shortcut. Similarly, we can also define auto-closing structures such as parentheses or brackets through the \lstinline{autoClosingPairs} attribute, which allows VS Code to automatically place closing parentheses or brackets. This is particularly useful for snail, as more opening parentheses or brackets require a closing partner. By defining \lstinline{autoClosingPairs}, a developer can expend less mental energy ensuring that each parenthesis and bracket has a closing pair, and more mental energy on developing code. The \lstinline{language-configuration.json} file for Snail Language Support can be found in appendix \ref{app:language-configuration}. For more attributes and configuration options, see the VS Code documentation on language configurations \cite{Microsoft_2023f}.

@@

Snail Language Support provides autocompletion suppport for common structures such as if-else clauses, while loops, and class definitions. This autocompletion is implemented through VS Code snippets. A snippet is a commonly used code structure that is defined by a \lstinline{prefix} (or a few options for a \lstinline{prefix}) and a \lstinline{body}. When a user types a snippet \lstinline{prefix}, they are given the option to replace the keyword with the \lstinline{body} of the snippet to complete the structure. Snippet bodies can also contain placeholders, which a user can visit in sequence while pressing their \lstinline{tab} button. For a full listing of the snippets that Snail Language Support provides, see appendix \ref{app:snippets-json}. Similarly, for more information on snippets in VS Code, see the documentation \cite{Microsoft_2023g}.

@@

Snail Language Support also supports syntax highlighting for the snail language as shown in figure \ref{fig:syntax-highlight}. VS Code's syntax highlighting makes heavy use of TextMate language grammars. A generic language grammar is a set of rules that govern what constitutes a valid statement for a given programming language \cite{pgrandinetti_2019}. A TextMate grammar is a specific format for defining a language grammar \cite{MacroMates_2021a}. TextMate grammars define patterns that match special elements of a text document and assign scope names to these elements. For example, keywords of a particular language might be assigned the \lstinline{keyword.control} scope name. These patterns are defined with regular expressions.

@@

TextMate grammars make use of the advanced Oniguruma regular expression syntax, developed by K. Kosako \cite{Microsoft_2023h, MacroMates_2021b}. The Oniguruma regular expression syntax offers advanced features such as case insentivity options, character groupings, quantifiers, and anchors.

@@

The engine tries matching text in the order we define our patterns. This means that the first pattern we define in \lstinline{patterns} is the first regular expression that the tokenization engine tries to match to a line in a snail file. It is important to pay attention to the order that patterns are defined. For example, Snail Language Support defines comments first, which ensures that all snail text, no matter how complicated or syntactically complex, is highlighted as a simple comment.\footnote{While defining our regular expression patterns, we found the Rubular program essesntial. It allowed us to iterate and test regular expressions more rapidly. We strongly recommend investigating this site while developing a TextMate grammar of your own: \url{https://rubular.com/}}

TextMate grammars also allow you to nest patterns inside of one another. This is essential for snail, as it allows the tokenization engine to match text inside of parentheses or brackets. TextMate grammars' ability to recursively reference the grammar itself (through the \lstinline{$self} keyword) is also essential for this functioanlity.

@@

Snail Language Support also provides realtime lexing and parsing error diagnostics when developing code in snail. To do this, Snail Language Support uses a language server that communicates via VS Code's langauge server protocol (LSP). We now discuss what a language server is, how it relates to the LSP, and how Snail Language Support uses langauge servers.

@@

A language server is a tool used to provide language-specific editing features such as advanced autocomplete, go-to definition, or automatic error checking \cite{Microsoft_2022a}. Many IDEs today utilize language servers by launching a language server process that runs in the background while a developer writes code. The development tool then sends a \emph{requests} to the language server for, and the language server \emph{responds} with the information requested by the development tool. The development tool then displays the information to the developer. This process is displayed in figure \ref{fig:language-server}.

Many language servers are implemented in the language they assist with \cite{Microsoft_2023j}. For example, a typical Python language server would be implemented in Python. This makes integrating language servers into development tools difficult, as many development tools are written in languages other than the language they are meant to develop. Further, a single language server might have to interact with a number of different development tools. For example, a Python language server might have to interact with VS Code, PyCharm, and Spyder. Without a standard framework for communication between development tools and language servers, development tools and language servers must implement specific functionality for each pairing. This problem is displayed in figure \ref{fig:no-lsp-theory}.

@@

\begin{figure}[t]
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{screenshots/dev-no-diagnostics-light.png}
            \subcaption{
                \centering
                    Developing snail code without realtime error diagnostics can make it difficult to spot syntax errors.
                }
            \label{fig:dev-no-diagnostics}
        \end{center}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{screenshots/dev-with-diagnostics-light.png}
            \subcaption{
                \centering
                    With Snail Language Support, a developer can spot easily spot syntax errors and fix them before runtime.
                }
            \label{fig:dev-with-diagnostics}
        \end{center}
    \end{subfigure}
    \caption{
        \centering
            Demonstrating the benefit of realtime error diagnostics provided in Snail Language Support.
        }
    \label{fig:dev-diagnostics}
\end{figure}

\begin{wrapfigure}{R}{0.5\textwidth}
    \includegraphics[width=0.5\textwidth]{png/lsp-sequence.png}
    \caption{
        \centering
            Modelling the interactions between a developer, a VS Code client, and snail language server while a developer opens and modifies a snail file.
    }
    \label{fig:lsp-sequence}
\end{wrapfigure}

@@

The snail-specific features are provided by modifying the snail interpreter to act as a language server. By passing a \lstinline{-s} or \lstinline{--server} flag while executing a snail file (i.e. \lstinline{snail -s hello-world.sl}), the snail interpreter outputs relevant error diagnostics including program exit status, error source, and error location.\footnote{Further modifying the snail interpreter to handle a variety of potential language server requests (not just requests for error diagnostics) is a great area for future work.} The \lstinline{server.ts} file handles this output and provides it to the VS Code client in the LSP format.

@@

Debugging tools are an integral part of the software development process. Modern debuggers are often implemented with a user-friendly UI to help developers debug their programs \cite{Microsoft_2022b,Matloff_Salzman_2008,Sinha_2017}. Many of these debuggers are built intertwined with their corresponding code editor's UI, which can vary from code editor to code editor. As a result, valuable debugging logic is intermingled with code-editor-specific UI logic. This makes it nearly impossible to implement a debugger that is capable of communicating with different code editors \cite{Microsoft_2021a}. This problem is illustrated in figure \ref{fig:no-dap-theory}.

Instead of asking a development tool to interface directly with a debugger, suppose a development tool interfaces with a debug adapter that translates a development tool's requests and sends them to a language-specific (but tool-agnostic) debugger. If the communication between the development tool and debug adapter is standardized, a language-specific debug adapter can be reused across multiple development tools.

@@

VS Code's DAP relies on requests and responses very similarly to the LSP \cite{Microsoft_2021b}. A request is a message that a development tool sends to a debug adapter, asking the debugger to perform some debugging task. The debug adapter communicates with the language-specific debugger to perform this operation and sends a corresponding response message to the development tool. The debug adapter \textbf{must} send a corresponding response after recieving a request, even if the response is an empty message. The DAP also defines reverse requests, which function the same as regular requests except that they are sent from the debug adapter to the development tool.

@@

Snail Language Support provides a debug adapter for the snail language in two parts, as shown in figure \ref{fig:dap-high-level}. First, an extension-level debug adapter is defined within Snail Language Support in the \lstinline{debugAdapter.ts} file, as found in appendix \ref{app:debug-adapter-ts}. This debug adapter is launched using the \lstinline{node.js} runtime within the VS Code client. The VS Code client sends DAP requests to the extension-level debug adapter via standard input (\lstinline{stdin}), and the debug adapter uses the transmission control protocol (TCP) to forward these DAP requests along to snail. Think of the extension-level debug adapter as a tool to echo DAP requests from VS Code along to snail.

@@

Next, the DAP requests are processed in a snail-level debug adapter that is defined as an additional module alongside the snail interpreter. This snail-level debug adapter is defined in \lstinline{snaildap.re} as shown in appendix \ref{app:snaildap} and is responsible for responding to DAP requests with the necessary snail-specific information.

\subsubsection{The \lstinline{snaildap} Snail Debug Adapter}

>>>>

    A high-quality programming environment (often an integrated development environment, or IDE) can be vital to enhancing developer productivity. Visual Studio Code (VS Code) is a popular, open-source text editor maintained by Microsoft. VS Code delivers language-specific features through freely dowloadable, community-built extensions on an online marketplace. Many of these extensions allow developers to take advantage of editing features such as syntax highlighting, code autocompletion, or debugging support. The snail language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming langauge meant to be implemented in a one-semester undergraduate course. We present a new VS Code extension to provide language support for the snail language. The extension implements support for syntax highlighting, rudimentary autocompletion, and static error-checking diagnostics using VS Code's Langauge Server Protocol. This report summarizes the contents of a VS Code extension and gives an overview of how an extension runs, particularly highlighting the functions of VS Code's Langauge Server Protocol. I also discuss how this extension can be futher developed to make use of VS Code's Debug Adapter Protocol to implement a debugger with breakpoints, start/stop behavior, and variable inspection.

@@

Much of software development in the present day takes place in integrated development environments (IDEs) \cite{JetBrains_2019}. An IDE is a collection of software development tools, such as a code editor, debugger, and build system, often unified under a similar user interface, with the goal of simplifying the software development process and enhancing developer productivity \cite{Gillis_Silverthorne_2018,Shyniaieva_2023}. In addition to composing these tools together, many IDEs also offer advanced features within their code editors such as syntax highlighting, code autocompletion, and error-checking diagnostics. Today, there are any number of different IDEs available for use with any given programming language, many offered as freely-downloadable for public use. For example, a developer looking to write code in Java may choose to do so in Eclipse, IntelliJ IDEA, or BlueJ \cite{EclipseFoundation_2023, JetBrains_2023,KingsProgrammingEducationToolsGroup_2022}. With so many high-quality IDEs available today, it is no surprise that 75\% of software developers today use an IDE in their everyday work \cite{JetBrains_2019}. Clearly, the IDE has become an integral part of the software development process today \cite{Vaniukov_2023}.

Visual Studio Code (VS Code), is a popular, open source text editor maintained by Microsoft \cite{StackOverflow_2022,Microsoft_2023a}. When first downloaded, VS Code is a lightweight text editor with minimal features. However, a number of community-built, freely downloadable extensions offer advanced langauge features. These extensions can be dowloaded on VS Code's online extension marketplace, and can turn VS Code into a very robust and powerful development environment for any programming task or language \cite{Microsoft_2023b}.

The snail programming language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming language meant to be implemented in a one-semester undergraduate course \cite{Angstadt_2023a}. In order to be implemented in a short time frame, snail is defined by limited features and a relatively annoying syntax. For example, the language lacks a for-loop structure, opting instead to provide only while-loops. Each if-statment \emph{requires} an else clause, even when the developer does not need to use one. Every statement must end with a semi-colon, which is not a huge issue. Unless a developer forgets to include a line-ending semi-colon, where they will discover that the resulting parse error message is wildly uninformative. While this design makes snail easier to implement, it makes it hard for a developer to write programs in snail.

@@

This report presents the Snail Language Support VS Code extension, which seeks to address the lack of programming support tools for the snail programming language. Snail Langauge Support provides several important features to make programming in the snail language easier. First, it features syntax highlighting to make reading snail code easier and help highlight key structures or keywords of the language. Further, it features rudimentary autocompletion with auto-closing brackets, braces, and quotes, as well as if-else, while-loop, and class definition snippets, reducing the burden of memorizing snail's strict and unintuitive syntax. The extension also has automatic, real-time error checking diagnostics that allow a user to see syntax or parse errors in a snail program before running it for themselves. Finally, the extension is structured to support a full debugger with breakpoints, step-in, step-over, and step-out functionality.

This paper will outline the process of building the Snail Language Support VS Code extension.\footnote{This paper is formatted using \LaTeX. Recognizing that using \LaTeX \hspace{1pt} is really hard, we make the \LaTeX \hspace{1pt} source code for this paper available on GitHub: \url{https://github.com/CharlesReinhardt/snail-language-support-paper/}} Specificaly, we will introduce the structure of a VS Code extension that is meant to add support for a new programming langauge. We will also discuss how this extension uses VS Code's language server protocol (LSP) to provide real-time error diagnostics \cite{Microsoft_2022a}. We will address how the Snail Language Support extension may be further developed to include debugging support with breakpoints, step in/out behavior, and variable inspection, particularly highlighting the role of VS Code's debug adapter protocol (DAP) \cite{Microsoft_2021a}. Finally, we will review good software development practices such as version control and documentation.

@@

While early IDEs certainly had their merits, they usually only supported one programming language. Microsoft's Visual Studio, released in 1997, was one of the first IDEs to package support for a variety of programming languages in one piece of software \cite{Microsoft_1997}. Visual Studio also featured extensive tools to aid in development for software on the early internet.

Today, we see a trend towards open source IDEs such as Eclipse, NetBeans, or VS Code \cite{Apache_2023, EclipseFoundation_2023, Microsoft_2023a}. Even open source editors such as these include intelligent code editors, debugging tools, and version control integration. Many of these editors have advanced features within their code editors, providing functions such syntax highlighting, code autocompletion, and real-time error diagnostics.

There are many benefits to developing software using modern IDEs. By bundling code editing, build systems, and program execution into one tool, modern IDEs reduce the time and effort a developer needs to put forth in order to write and test a piece of code \cite{Gillis_Silverthorne_2018}. This also reduces the number of decisions a developer has to make while developing code, which can help increase productivity. Using an IDE can also standardize the software development process, by either helping a group of people use a consistent UI (and know how to help eachother), or allow a developer to avoid switching applications to complete a single task \cite{Veracode_2020}.

@@

Debugging is the process of searching for and fixing unexpected errors in a piece of code \cite{AWS_2023}. Early techniques of debugging software involved physically printing machine output and reading through it step-by-step, searching for a potential error \cite{RevDeBug_2020}. Some computer systems, such as the IBM 704 Data Processing system released in 1968, allowed programmers to print information stored in specified memory locations in specified forms to assist with debugging activities \cite{MIT_1957, IBMArchives_2023}.

@@

Visual Studio Code (VS Code) is a popular, open source code editor. VS Code is designed to be fast and lightweight, with a focus on allowing a developer to write, test, and debug source code quickly \cite{Microsoft_2023c}. To achieve this, VS Code uses native, web, and language-specific technologies. Tools like Electron allow VS Code to run on multiple platforms and operating systems using common web technologies like JavaScript, HTML, and CSS \cite{Electron_2023}.

@@

Most VS Code extensions share a similar directory structure. These structures may vary depending on what purpose the extension is meant to serve. In figure \ref{fig:directory-structure}, we see a diagram of a typical language support VS Code extension.\footnote{The exact directory structure of the Snail Language Support extension is not included in this paper. For a deeper look at the directory structure and source code for Snail Language Support, see the git repository hosted on GitHub: \url{https://github.com/snail-language/snail-language-support}} First, the \lstinline{.vscode} directory configures how we run and test our exstension locally. Through the \lstinline{launch.json} and \lstinline{tasks.json} files, we define various tasks and configurations to streamline the build and testing process.\footnote{From personal experience, we strongly recommend taking the time to understand these files before developing a VS Code extension. It will likely save painstaking debugging in the near future. See launch configurations: \url{https://code.visualstudio.com/docs/editor/debugging\#_launch-configurations} and tasks: \url{https://code.visualstudio.com/docs/editor/tasks}} Note that these configuration files do not get packaged with a released VS Code extension. Next, a \lstinline{README.md} file allows us to advertise the features of our extension. This file is rendered in the VS Code extension marketplace and is meant to be a user's first impression of an extension.

@@

The \lstinline{src} directory contains the aforementioned TypeScript source code that runs our extension. In this directory, the \lstinline{extension.ts} file defines special actions for our extension and launches the extension within an existing VS Code window. The \lstinline{server.ts} file defines a language server that allows our extension to take advantage of real-time error diagnostics. We further discuss language servers later in this paper.

@@

When a VS Code window launches, it also launches all extensions the user has downloaded and enabled by calling each extension's \lstinline{activate} function, defined in the extension's source code. In Snail Language Support's case, the \lstinline{activate} function is defined in the \lstinline{extension.ts} file. This file has access to the VS Code extension API, allowing a developer to define certain VS Code client UI updates \cite{Microsoft_2023d}. For example, Snail Language Support can display a VS Code style error message to the user if the user does not have a compatibale version of the snail language dowloaded.

The \lstinline{activate} function is responsible for a few different behaviors. First, the \lstinline{activate} function registers multiple command handlers to watch for commands input from a user. A command handler is an example of the observer design pattern, which allows a developer to address a variety of potential input events that are experienced during runtime \cite{GangOfFour_1995}. A developer can register an observer that listens for a particular event and executes a block of code once that event is detected. Snail Language Support defines command handlers to detect when a user chooses to run or debug a snail file. We will touch on the topic of debugging within Snail Language Support more later.

@@

Once the extension setup inside the \lstinline{activate} function is complete, VS Code enables the extension within the workspace, allowing a developer to utilize the functions of the extension. Once the VS Code window is closed, or the extension is manually disabled, VS Code calls the \lstinline{deactivate} function, which shuts down any separate processes left over from the extension run.

To inspect Snail Language Support's \lstinline{extension.ts} file and \lstinline{activate} function more closely, see appendix \ref{app:extension-ts}

@@

First, we can define some metadata for our extension, such as a name, author, description, and version. We can also define which version of the VS Code engine this extension expects. This ensures that an extension knows which VS Code APIs it has proper access to. For example, an extension that expects a VS Code version of \lstinline{1.7.0} won't be able to take advantage of an API released in \lstinline{1.8.0} \cite{Microsoft_2023e}. We can also define some categories that our extension falls under, helping our extension gain visibility on the online VS Code marketplace.

@@

Snail Language Support also provides language configuration settings supported by a VS Code extension. These configurations are found in \lstinline{language-configuration.json} file. We can define what is considered a comment through the \lstinline{comments.lineComment} and \lstinline{comments.blockComment}attributes, which allows a user to comment lines using VS Code's comment shortcut. Similarly, we can also define auto-closing structures such as parentheses or brackets through the \lstinline{autoClosingPairs} attribute, which allows VS Code to automatically place closing parentheses or brackets. This is particularly useful for snail, as most opening parentheses or brackets require a closing partner. By defining \lstinline{autoClosingPairs}, a developer can expend less mental energy ensuring that each parenthesis and bracket has a closing pair, and more mental energy on developing code. The \lstinline{language-configuration.json} file for Snail Language Support can be found in appendix \ref{app:language-configuration}. For more attributes and configuration options, see the VS Code documentation on language configurations \cite{Microsoft_2023f}.

@@

Snail Language Support provides autocompletion suppport for common structures such as if-else clauses, while-loops, and class definitions. This autocompletion is implemented through VS Code snippets. A snippet is a commonly used code structure that is defined by a \lstinline{prefix} (or a few options for a \lstinline{prefix}) and a \lstinline{body}. When a user types a snippet \lstinline{prefix}, they are given the option to replace the keyword with the \lstinline{body} of the snippet to complete the structure. Snippet bodies can also contain placeholders, which a user can visit in sequence while pressing their \lstinline{tab} button. For a full listing of the snippets that Snail Language Support provides, see appendix \ref{app:snippets-json}. Similarly, for more information on snippets in VS Code, see the documentation \cite{Microsoft_2023g}.

@@

Snail Language Support also supports syntax highlighting for the snail language as shown in figure \ref{fig:syntax-highlight}. VS Code's syntax highlighting makes use of TextMate language grammars. A generic language grammar is a set of rules that govern what constitutes a valid statement for a given programming language \cite{pgrandinetti_2019}. A TextMate grammar is a specific format for defining a language grammar \cite{MacroMates_2021a}. TextMate grammars define patterns that match special elements of a text document and assign scope names to these elements. For example, keywords of a particular language might be assigned the \lstinline{keyword.control} scope name. These patterns are defined with regular expressions.

@@

TextMate grammars use the advanced Oniguruma regular expression syntax, developed by K. Kosako \cite{Microsoft_2023h, MacroMates_2021b}. The Oniguruma regular expression syntax offers advanced features such as case insentivity options, character groupings, quantifiers, and anchors.

@@

The engine tries matching text in the order we define our patterns. This means that the first pattern we define in \lstinline{patterns} is the first regular expression that the tokenization engine tries to match to a line in a snail file. It is important to pay attention to the order that patterns are defined. For example, Snail Language Support defines comments first, which ensures that all comments in snail, no matter how complicated or syntactically complex, is highlighted as a simple comment.\footnote{While defining our regular expression patterns, we found the Rubular program essential. It allowed us to iterate and test regular expressions more rapidly. We strongly recommend investigating this site while developing a TextMate grammar of your own: \url{https://rubular.com/}}

TextMate grammars also allow you to nest patterns inside of one another. This is essential for snail, as it allows the tokenization engine to match text inside of parentheses or brackets. TextMate grammars' ability to recursively reference the grammar itself (through the \lstinline{$self} keyword) proves essential for this functioanlity.

@@

Snail Language Support also provides real-time lexing and parsing error diagnostics when developing code in snail. To do this, Snail Language Support uses a language server that communicates via VS Code's langauge server protocol (LSP). We now discuss what a language server is, how it relates to the LSP, and how Snail Language Support uses langauge servers.

@@

A language server is a tool used to provide language-specific editing features such as advanced autocomplete, go-to definition, or automatic error checking \cite{Microsoft_2022a}. Many IDEs today utilize language servers by launching a language server process that runs in the background while a developer writes code. The development tool then sends \emph{requests} to the language server and the language server \emph{responds} with the information requested by the development tool. The development tool then displays the information to the developer. This process is displayed in figure \ref{fig:language-server}.

Many language servers are implemented in the language they emulate \cite{Microsoft_2023j}. For example, a typical Python language server would be implemented in Python. This makes integrating language servers into development tools difficult. First, most development tools for specific languages are not implemented in those specific languages. For example, most development tools for Python are not implemented with Python. Further, a single language server might have to interact with a number of different development tools. For example, a Python language server might have to interact with VS Code, PyCharm, and Spyder. Without a standard framework for communication between development tools and language servers, development tools and language servers must implement specific functionality for each pairing. This problem is displayed in figure \ref{fig:no-lsp-theory}.

@@

\begin{figure}[t]
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{screenshots/dev-no-diagnostics-light.png}
            \subcaption{
                \centering
                    Developing snail code without real-time error diagnostics can make it difficult to spot syntax errors.
                }
            \label{fig:dev-no-diagnostics}
        \end{center}
    \end{subfigure}
    \hspace{0.05\textwidth}
    \begin{subfigure}{0.45\textwidth}
        \begin{center}
            \includegraphics[width=\textwidth]{screenshots/dev-with-diagnostics-light.png}
            \subcaption{
                \centering
                    With Snail Language Support, a developer can spot easily spot syntax errors and fix them before runtime.
                }
            \label{fig:dev-with-diagnostics}
        \end{center}
    \end{subfigure}
    \caption{
        \centering
            Demonstrating the benefit of real-time error diagnostics provided in Snail Language Support.
        }
    \label{fig:dev-diagnostics}
\end{figure}

@@

The snail-specific features are provided by modifying the snail interpreter to act as a language server. By passing the \lstinline{-s} or \lstinline{--server} flag while executing a snail file (i.e. \lstinline{snail -s hello-world.sl}), the snail interpreter outputs relevant error diagnostics including program exit status, error source, and error location.\footnote{Further modifying the snail interpreter to handle a variety of potential language server requests (not just requests for error diagnostics) is a great area for future work.} The \lstinline{server.ts} file handles this output and provides it to the VS Code client in the LSP format.

\begin{wrapfigure}{R}{0.5\textwidth}
    \includegraphics[width=0.5\textwidth]{png/lsp-sequence.png}
    \caption{
        \centering
            Modelling the interactions between a developer, VS Code client, and snail language server while a developer modifies a snail file.
    }
    \label{fig:lsp-sequence}
\end{wrapfigure}

@@

Debugging tools are an integral part of the software development process. Modern debuggers are often implemented with a user-friendly UI to help developers debug their programs \cite{Microsoft_2022b,Matloff_Salzman_2008,Sinha_2017}. Many of these debuggers are built intertwined with their corresponding code editor's UI, which can vary from code editor to code editor. As a result, valuable debugging logic is intermingled with editor-specific UI logic. This makes it nearly impossible to implement a debugger that is capable of communicating with different code editors \cite{Microsoft_2021a}. This problem is illustrated in figure \ref{fig:no-dap-theory}.

Instead of asking a development tool to interface directly with a debugger, suppose a development tool interfaces with a debug adapter that translates a development tool's requests and sends them to a language-specific (but tool-agnostic) debugger. If the communication between the development tool and debug adapter is standardized, language-specific debuggers and debug adapters can be reused across multiple development tools.

@@

VS Code's DAP relies on requests and responses very similarly to the LSP \cite{Microsoft_2021b}. A request is a message that a development tool sends to a debug adapter, asking the debugger to perform some debugging task. The debug adapter communicates with the language-specific debugger to perform this operation and sends a corresponding response message back to the development tool. The debug adapter \textbf{must} send a corresponding response after recieving a request, even if the response is an empty message. The DAP also defines reverse requests, which function the same as regular requests except that they are sent from the debug adapter to the development tool.

@@

@@

Snail Language Support provides a debug adapter for the snail language in two parts, as shown in figure \ref{fig:dap-high-level}. First, we define an extension-level debug adapter within Snail Language Support in the \lstinline{debugAdapter.ts} file, as found in appendix \ref{app:debug-adapter-ts}. This debug adapter is launched using the \lstinline{node.js} runtime within the VS Code client. The VS Code client sends DAP requests to the extension-level debug adapter via standard input (\lstinline{stdin}), and the debug adapter uses the transmission control protocol (TCP) to forward these DAP requests along to snail. Think of the extension-level debug adapter as a tool to echo DAP requests from VS Code along to snail.

Next, the DAP requests are processed in a snail-level debug adapter that we define as an additional module alongside the snail interpreter in the \lstinline{snaildap.re} file as shown in appendix \ref{app:snaildap}. The snail-level debug adapter is responsible for responding to DAP requests with the necessary snail-specific information.
